Hypothesis tests for multiple responses regression models in \proglang{R}: The \pkg{htmcglm} Package

### 2022-07-25 5_artigos/journal-of-statistical-software/draft/draft2/draft-jss-lineu.Rnw
atualiza rascunhos e dissertacao

\title{Hypothesis tests for multivariate regression with non-Gaussian data in \proglang{R}: The \pkg{htmcglm} Package}

\Plaintitle{Hypothesis tests for multivariate regression with non-Gaussian data in R: The htmcglm Package}

\Shorttitle{htmcglm: hypothesis tests for multivariate regression}

@@

This article describes the R package htmcglm implemented for performing hypothesis tests on regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs) fiited by package mcglm. Our implementations are based on the proposal to use the Wald test to perform general hypothesis tests, ANOVAs, MANOVAs and multiple comparison tests. McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. Our contribution is aimed at providing tools for a better interpretation of the estimated parameters and thus extracting more information and conclusions about the problems modeled through the class. By studying the regression parameters, it is possible to assess the effect of the explanatory variable(s) on the response(s). By studying the dispersion parameters, the effect of the correlation between study units can be evaluated, very useful in situations where the observations of the data set are correlated with each other, such as in longitudinal, temporal and repeated measures studies. Our implementations provide a procedural and safe way to answer common questions in the context of multivariate modeling, such as: which variables are associated with the outcome of the phenomenon of interest, if there is an effect of the correlation structure between individuals in the study, if the effect of a certain variable is the same regardless of the response, among others. Illustrations in this article cover the most common hypothesis tests of interest in practical contexts in modeling multivariate regression problems.

@@

The \pkg{htmcglm} package for \proglang{R} \citep{R2022} provides functions for performing hypothesis testing on parameters of multivariate covariance generalized linear models (McGLMs; \citet{Bonat16}) fitted using the \pkg{mcglm} package \citep{mcglm}.

McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. When we work on the McGLMs class we estimate parameters of regression, dispersion, power and correlation. Each set of parameters has a very useful practical interpretation.

By studying the regression parameters, it is possible to assess the effect of the explanatory variable(s) on the response(s). By studying the dispersion parameters, the effect of the correlation between study units can be evaluated, very useful in situations where the observations of the data set are correlated with each other, such as in longitudinal, temporal and repeated measures studies. The power parameters give us an indication of which probability distribution best fits the problem. And correlation parameters estimate the strength of association between responses in a multivariate problem.

The development of hypothesis tests for the purpose of evaluating these quantities is of great value in practical problems and leads to procedural forms for evaluating the resulting quantities of the model. The \pkg{htmcglm} package is a full \proglang{R} implementation with functions based on Wald statistics to evaluate regression and dispersion parameters. The features include functions for general linear hypothesis testing, univariate and multivariate analysis of variance tables, as well as multiple comparison tests.

Package \pkg{htmcglm} is available on github at https://github.com/lineu96/htmcglm and complement the functions available in the package \pkg{mcglm}.

@@

However, when dealing with multivariate covariance generalized linear models fitted in the \pkg{mcglm} package, there is only one type of analysis of variance implemented in the library and there are no options for performing general linear hypothesis tests, nor multiple comparison tests. Therefore, as it is a flexible class of models with high power of application to practical problems, our general objective is to provide implementations that allow performing hypothesis tests for McGLMs in such a way that it is possible to test general linear hypotheses, generate analysis of variance tables, multivariate analysis of variance tables and multiple comparisons tests.

The article is organized as follows. In \autoref{sec:mcglm} we present a review of the general structure and estimation of the parameters of a McGLM, based on the ideas of \citet{Bonat16}. In \autoref{sec:wald} the details of the Wald test to evaluate assumptions about parameters of a McGLM are presented. In \autoref{sec:implementacao} we introduce the \proglang{R} implementation discussing the main functions available in the \pkg{htmcglm} package. We illustrate the package usage through some examples in \autoref{sec:exemplos}. Finally, \autoref{sec:conclusao} presents a discussion and directions for future work on the improvement of the \pkg{htmcglm} package.

@@

Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ a matrix of response variables and $\boldsymbol{M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$, is denoted by $\Sigma_r$ and has dimension $N \times N$. In addition, consider a $R \times R$ correlation matrix, denoted by $\Sigma_b$, to describe the correlation between the response variables. The McGLMs \citep{Bonat16} are defined by:

@@

\noindent where the functions $g_r()$ are the traditional link functions; $\boldsymbol{X}_r$ denotes a design matrix $N \times k_r$; $\boldsymbol{\beta}_r$ denotes a vector $k_r \times 1$ of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{\Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ is the Generalized Kronecker product \citep{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. the operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

For continuous, binary, binomial, proportions or indices, the variance and covariance matrix $\boldsymbol{\Sigma}_r$ is given by:

@@

In the case of response variables that are counts, the variance and covariance matrix for each response variable is given by:

@@

\noindent where $\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right) = diag(\vartheta(\boldsymbol{\mu}_r; p_r))$ denotes a diagonal matrix in which the entries are given by the variance function $\vartheta(\cdot; p_r)$ applied to vector elements $\boldsymbol{\mu}_r$. Different choices of variance functions $\vartheta(\cdot; p_r)$ imply different assumptions about the distribution of the response variable. We will mention 3 options of variance functions: the power variance function, the Poisson-Tweedie dispersion function and the binomial variance function.

The power variance function characterizes the Tweedie family of distributions, is given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_r}_r$, in which some distributions stand out: Normal ($p$ = 0), Poisson ($p$ = 1), gama ($p$ = 2) and  inverse Gaussian ($p$ = 3) \citep{Jorgensen87, Jorgensen97}.

The Poisson-Tweedie dispersion function \citep{Jorgensen15} is indicated for events defined by counts. Is given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. We thus have a rich class of models for dealing with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

Finally, the binomial variance function, given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_{r1}}_r(1 - \mu_r)^{p_{r2}}$ is indicated when the response variable is binary, restricted to an interval or when there is a number of successes in a number of trials.

It is possible to notice that the power parameter $p$ appears in all the variance functions discussed. This parameter is especially important because it is an index that distinguishes different probability distributions that are important in the modeling context and, for this reason, can be used as a tool for automatic selection of the probability distribution that best suits the problem.

@@

\noindent where $h()$ is the covariance link function, $Z_{rd}$ with $d$ = 0,$\ldots$, D are matrices that represent the covariance structure present in each response variable $r$ and $\boldsymbol{\tau_r}$ = $(\tau_{r0}, \ldots, \tau_{rD})$ is a vector $(D + 1) \times 1$ of dispersion parameters.

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by \citet{Pinheiro96} and it is possible to select combinations of matrices to obtain the best known models in the literature for longitudinal data, time series, spatial and spatio-temporal data. Further details are discussed by \citet{Demidenko13}.

In this way, the McGLMs configure a general framework for analysis via regression models for non-Gaussian data with multiple responses, in which no assumptions are made regarding the independence of the observations. The class is defined by 3 functions (link, variance and covariance), in addition to a linear predictor and a matrix linear predictor for each response under analysis.

@@

McGLMs are fitted based on the estimating function approach described in detail by \citet{Bonat16} and \citet{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

McGLM's second-moment assumptions allow the division of parameters into two subsets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^{\top}$. This way, $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a vector $K \times 1$ of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top, \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a vector $Q \times 1$ of dispersion parameters. Furthermore, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the stacked vector $NR \times 1$ of the matrix of response variables $\boldsymbol{Y}_{N \times R}$ and $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^\top)^\top$ denotes the stacked vector $NR \times 1$ of the matrix of expected values $\boldsymbol{M}_{N \times R}$.

To estimate the regression parameters, the quasi-score function \citep{Liang86} is used, represented by

@@

\noindent where $\boldsymbol{D} = \nabla_{\boldsymbol{\beta}} \mathcal{M}$ is a matrix $NR \times K$, and $\nabla_{\boldsymbol{\beta}}$ denotes the gradient operator. Using the quasi-score function the sensitivity matrix $K \times K$ of $\psi_{\boldsymbol{\beta}}$ is given by

@@

For the dispersion parameters, the Pearson estimating function is used, defined in the form

@@

\noindent The entry $(i,j)$ of the variability matrix $Q \times Q$ of $\psi_{\boldsymbol{\lambda}}$ is defined by

@@

\noindent where $k_{l}^{(4)}$ denotes the fourth cumulant of $\mathcal{Y}_{l}$. In the McGLM estimation process, empirical versions are used.

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, \citet{Bonat16} obtained the cross-sensitivity and variability matrices, denoted by $S_{\boldsymbol{\lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details in \citet{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

@@

Let $\boldsymbol{\hat{\theta}} = (\boldsymbol{\hat{\beta}^{\top}}, \boldsymbol{\hat{\lambda}^{\top}})^{\top}$ the estimator based on estimating functions of $\boldsymbol{\theta}$. Then, the asymptotic distribution of $\boldsymbol{\hat{\theta}}$ is

@@

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$ the modified Chaser algorithm is used, proposed by \citet{jorg04}, which is defined as

$$
\begin{aligned}
\begin{matrix}
\boldsymbol{\beta}^{(i+1)} = \boldsymbol{\beta}^{(i)}- S_{\boldsymbol{\beta}}^{-1} \psi \boldsymbol{\beta} (\boldsymbol{\beta}^{(i)}, \boldsymbol{\lambda}^{(i)}), \\ 
\boldsymbol{\lambda}^{(i+1)} = \boldsymbol{\lambda}^{(i)}\alpha S_{\boldsymbol{\lambda}}^{-1} \psi \boldsymbol{\lambda} (\boldsymbol{\beta}^{(i+1)}, \boldsymbol{\lambda}^{(i)}).
\end{matrix}
\end{aligned}
$$

@@

Following the ideas of \textbf{REF MEU ARTIGO}, consider $\boldsymbol{\theta^{*}}$ the vector $h \times 1$ of parameters disregarding the correlation parameters, ie, $\boldsymbol{\theta^{*}}$ only refers to regression, dispersion or power parameters. The estimates of the parameters of $\boldsymbol{\theta^{*}}$ are given by $\boldsymbol{\hat\theta^{*}}$. In a similar way, consider $J^{\boldsymbol{*}-1}$ the inverse of the Godambe information matrix disregarding the correlation parameters, of dimension $h \times h$. Let $\boldsymbol{L}$ a specification matrix of hypotheses to be tested, of dimension $s \times h$ and $\boldsymbol{c}$ a vector of dimension $s \times 1$ with the values under the null hypothesis, where $s$ denotes the number of restrictions. The hypotheses to be tested can be written as:

@@

\noindent In this way, the generalization of the Wald test statistic to verify the validity of a hypothesis about parameters of a McGLM is given by:

@@

In general, each column of the matrix $\boldsymbol{L}$ corresponds to one of the $h$ parameters of $\boldsymbol{\theta^{*}}$ and each row to a constraint. Its construction basically consists of filling the matrix with 0, 1 and eventually -1 in such a way that the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ correctly represents the hypotheses of interest. The correct specification of $\boldsymbol{L}$ allows testing any parameter individually or even formulating hypotheses for several parameters.

\textbf{REF MEU ARTIGO} presents examples of how to test different types of hypotheses of interest that arise in practical contexts. We will present in this work two of these examples: hypotheses for multiple parameters and hypotheses about regression or dispersion parameters for responses under the same predictor.

For purposes of illustration, consider the situation in which you want to investigate whether a numeric variable $X_1$ has an effect on two response variables, denoted by $Y_1$ and $Y_2$. A bivariate McGLM for this problem may have a predictor given by:

@@

\noindent where the index $r$ denotes the response variable, r = 1,2; $\beta_{r0}$ represents the intercept; $\beta_{r1}$ a regression parameter associated with the variable $X_1$. Assume that each response has only one dispersion parameter $\tau_{r0}$ and that the power parameters were fixed. Therefore, it is a problem in which there are two response variables and only one explanatory variable. Assume that the units under study are independent, so $Z_0 = I$.

Suppose the interest is to assess whether there is sufficient evidence to state that there is an effect of the explanatory variable $X_1$ on both responses simultaneously. In this case we will have to test 2 parameters: $\beta_{11}$, which associates $X_1$ with the first response; and $\beta_{21}$, which associates $X_1$ with the second response. We can write the hypothesis as follows:

@@

The vector $\boldsymbol{\theta^{*}}$ has six elements and the matrix $\boldsymbol{L}$ has six columns. In this case we are testing two parameters, so the matrix $\boldsymbol{L}$ has two rows. These lines are composed of zeros, except in the columns referring to the parameter of interest. It is simple to verify that the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ represents the hypothesis of interest initially postulated in \autoref{eq:ex2}. Thus, the asymptotic distribution of the test is $\chi^2_2$.

@@

Suppose that, in this example, the hypotheses of interest are still written as in the form of \autoref{eq:ex2}. However, as this is a bivariate model with the same predictor for the two responses, the hypothesis of interest are the same between responses and involves only regression parameters, it is convenient to write the matrix $\boldsymbol{L}$ as the Kronecker product of two matrices: a matrix $\boldsymbol{G}$ and a matrix $\boldsymbol{F}$, ie, $\boldsymbol{L}$ = $\boldsymbol{G} \otimes \boldsymbol{F}$. In this way, the matrix $\boldsymbol{G}$ has dimension $R \times R$ and specifies the hypotheses about the responses, whereas the matrix $\boldsymbol{F}$ specifies the hypotheses between variables and has dimension ${s}' \times {h}'$, where ${s}'$ is the number of linear constraints, that is, the number of parameters tested for a single response, and ${h}'$ is the total number of coefficients of regression or dispersion of the response. Therefore, the matrix $\boldsymbol{L}$ has dimension (${s}'R \times h$).

In general, the matrix $\boldsymbol{G}$ is an identity matrix with a dimension equal to the number of responses analyzed in the model. Whereas the matrix $\boldsymbol{F}$ is equivalent to a matrix $\boldsymbol{L}$ if there was only a single response in the model and only regression or dispersion parameters. We use the Kronecker product of these two matrices to ensure that the hypothesis described in the $\boldsymbol{F}$ matrix will be tested on the $R$ model responses.

@@

Based on the proposed use of the Wald test for McGLMs, \textbf{REF MEU ARTIGO} proposed three different procedures for generating ANOVA and MANOVA tables for regression parameters, and a procedure similar to ANOVA and MANOVA to evaluate the dispersion parameters of a model. In the case of ANOVAs, a table is generated for each response variable. For MANOVAs only one table is generated, therefore, in order to be able to perform MANOVAs, the responses must be subject to the same predictor.

For purposes of illustration, consider the situation where the objective is to investigate whether two numeric variables denoted by $X_1$ and $X_2$ have an effect on two response variables denoted by $Y_1$ and $Y_2$. For this case, consider the following predictor:

@@

In this section we provide practical examples of using the functions implemented in the \pkg{htmcglm} package based on multivariate models fitted with the \pkg{mcglm} package.

>>>>

\title{Hypothesis tests for multiple responses regression models in \proglang{R}: The \pkg{htmcglm} Package}

\Plaintitle{Hypothesis tests for multiple responses regression models in R: The htmcglm Package}

\Shorttitle{htmcglm: hypothesis tests for multiple responses regression models}

@@

This article describes the \proglang{R} package \pkg{htmcglm} implemented for performing hypothesis tests on regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. The proposed package considers the Wald statistics to perform general hypothesis tests and build tailored ANOVAs, MANOVAs and multiple comparison tests. The goal of the package is to provide tools to improve the interpretation of regression and dispersion parameters. We assess the effects of the covariates on the response variables by testing the regression coefficients. Similarly, we perform tests on the dispersion coefficients in order to assess the correlation between study units. It could be of interest in situations where the data observations are correlated with each other, such as in longitudinal, times series, spatial and repeated measures studies. The \pkg{htmcglm} package provides a user friendly interface to perform MANOVA like tests as well as multivariate hypothesis tests for models of the mcglm class. We describe the package implementation and illustrate it through the analysis of two data sets. The first deals with an experiment on soybean yield; the problem has three response variables of different types (continuous, counting and binomial) and three explanatory variables (amount of water, fertilization and block). The second dataset addresses a problem where responses are longitudinal bivariate counts of hunting animals; the explanatory variables used are the hunting method and sex of the animal. With these examples we were able to illustrate several tests in which the proposal proves to be useful for the evaluation of regression and dispersion parameters both in problems with dependent or independent observations.

@@

The \pkg{htmcglm} package for \proglang{R} \citep{R2022} provides functions for performing hypothesis testing on parameters of multivariate covariance generalized linear models (McGLMs; \citet{Bonat16}) as fitted by the \pkg{mcglm} package \citep{mcglm}.

McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
McGLMS are specified by a set of regression, dispersion, power and correlation parameters. Each set of parameters has a very useful practical interpretation.

By analysing the regression parameters, it is possible to assess the effect of the explanatory variables on the response variables. 
Simililarly, by analysing the dispersion parameters, we can assess the correlation structure between study units.
It is useful in situations where the observations of the data set are correlated with each other, such as in longitudinal, times series and repeated measures studies. 
The power parameters provide us an indication of which probability distribution could fit well to the response variable. Finally, the correlation parameters measure 
the strength of the association between response variables in a multivariate context.

The development of hypothesis tests for the purpose of evaluating these quantities is of great interest in practical problems and leads to procedural forms for evaluating the resulting quantities of the model. 
The \pkg{htmcglm} package is a full \proglang{R} implementation with functions based on Wald statistics to evaluate regression and dispersion parameters. 
The features include functions for general linear hypothesis testing, univariate and multivariate analysis of variance tables, as well as multivariate multiple comparison tests.

The \pkg{htmcglm} package is available from the Comprehensive R Archive Network (CRAN) at \href{https://cran.r-project.org/package=htmcglm}{https://cran.r-project.org/package=htmcglm} and complement the functions available in the \pkg{mcglm} package \citep{mcglm}.

@@

However, when dealing with multivariate covariance generalized linear models fitted in the \pkg{mcglm} package, there is only one type of analysis of variance implemented in the library and there are no options for performing general linear hypothesis tests, nor multiple comparison tests. Therefore, as it is a flexible class of models with high application potential, our goal is to provide computational implementation of hypothesis tests for McGLMs in such a way that it is possible to test general linear hypotheses, generate analysis of variance tables, multivariate analysis of variance tables and multiple comparisons tests.

The article is organized as follows. In \autoref{sec:mcglm} we present a review of the general structure and estimation of the parameters of a McGLM, based on the ideas of \citet{Bonat16}. In \autoref{sec:wald} the details of the Wald test to evaluate assumptions about parameters of a McGLM are presented. In \autoref{sec:implementacao} we introduce the \proglang{R} implementation discussing the main functions available in the \pkg{htmcglm} package. We illustrate the package use through some examples in \autoref{sec:exemplos}. Finally, \autoref{sec:conclusao} presents a discussion and directions for future work on the improvement of the \pkg{htmcglm} package.

@@

In this section, we present a review of the McGLMs specification. Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ is a matrix of response variables and $\boldsymbol{M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ is a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$, is denoted by $\Sigma_r$ and has dimension $N \times N$. In addition, consider an $R \times R$ correlation matrix, denoted by $\Sigma_b$, to describe the correlation between the response variables. The McGLMs \citep{Bonat16} are specified by:

@@

\noindent where the functions $g_r()$ are standard link functions; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{\Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ is the Generalized Kronecker product \citep{martinez13}. The matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is a $N \times N$ identity matrix.

For continuous, binary, binomial and bounded data, the variance and covariance matrix $\boldsymbol{\Sigma}_r$ is given by:

@@

In the case of count response variables, the variance and covariance matrix for each response variable is given by:

@@

\noindent where $\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right) = diag(\vartheta(\boldsymbol{\mu}_r; p_r))$ denotes a diagonal matrix in which the entries are given by the variance function $\vartheta(\cdot; p_r)$ applied elementwise to the vector $\boldsymbol{\mu}_r$. Different choices of variance functions $\vartheta(\cdot; p_r)$ imply different assumptions about the distribution of the response variable. We mention three options of variance functions: the Tweedie variance function, the Poisson-Tweedie dispersion function and the binomial variance function.

The Tweedie variance function characterizes the Tweedie family of distributions, is given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_r}_r$, in which some distributions stand out: Normal ($p$ = 0), Poisson ($p$ = 1), gama ($p$ = 2) and inverse Gaussian ($p$ = 3) \citep{Jorgensen87, Jorgensen97}.

The Poisson-Tweedie dispersion function \citep{Jorgensen15} is indicated for events defined by counts. It is given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. Thus, we have a rich class of models for dealing with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

Finally, the binomial variance function, given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_{r1}}_r(1 - \mu_r)^{p_{r2}}$ can deal with binary, binomial and continuous bounded respose variables.
It is possible to notice that the power parameter $p$ appears in all the variance functions discussed. This parameter is important because it is an index that distinguishes between different 
probability distributions. Thus, it can be regarded as a tool for probability distribution automatic selection.

@@

\noindent where $h()$ is the covariance link function, $Z_{rd}$ with $d$ = 0,$\ldots$, $D$ are matrices that represent the covariance structure for each response variable $r$ and $\boldsymbol{\tau_r}$ = $(\tau_{r0}, \ldots, \tau_{rD})$ is a $(D + 1) \times 1$ vector of dispersion parameters.

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by \citet{Pinheiro96} and it is possible to select combinations of matrices to obtain the some well-known models in the literature for longitudinal data, time series, spatial and spatio-temporal data. Further details are discussed by \citet{Demidenko13}.

Thus, the McGLMs configure a general framework for analysis via regression models for Gaussian and non-Gaussian data with multiple responses, in which no assumptions are made regarding the independence of the observations. The class is defined by three functions (link, variance and covariance), in addition to a linear predictor and a matrix linear predictor for each response under analysis.

@@

McGLMs are fitted based on the estimating function approach described in detail by \citet{Bonat16} and \citet{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of the estimators based on estimating functions.

McGLM's second-moment assumptions allow us to split the parameter vector into two subsets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^{\top}$. Thus, $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a  $K \times 1$ vector of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top, \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a $Q \times 1$ vector of dispersion parameters. Furthermore, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the $NR \times 1$ stacked response variables vector. Similarly, $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^\top)^\top$ denotes the $NR \times 1$ stacked expected values vector.

In order to estimate the regression parameters, the quasi-score function \citep{Liang86} is given by

@@

\noindent where $\boldsymbol{D} = \nabla_{\boldsymbol{\beta}} \mathcal{M}$ is a $NR \times K$ matrix and $\nabla_{\boldsymbol{\beta}}$ denotes the gradient operator. The $K \times K$ sensitivity matrix of $\psi_{\boldsymbol{\beta}}$ is given by

@@

For the dispersion parameters, the following Pearson estimating function is adopted,

@@

\noindent The entry $(i,j)$ of the $Q \times Q$ variability matrix of $\psi_{\boldsymbol{\lambda}}$ is defined by

@@

\noindent where $k_{l}^{(4)}$ denotes the fourth cumulant of $\mathcal{Y}_{l}$. In the McGLM estimation process its empirical version is used.

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, \citet{Bonat16} obtained the cross-sensitivity and variability matrices, denoted by $S_{\boldsymbol{\lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details, see \citet{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

@@

Let $\boldsymbol{\hat{\theta}} = (\boldsymbol{\hat{\beta}^{\top}}, \boldsymbol{\hat{\lambda}^{\top}})^{\top}$ be the estimating functions estimators of $\boldsymbol{\theta}$. Then, the asymptotic distribution of $\boldsymbol{\hat{\theta}}$ is

@@

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$ the following modified Chaser algorithm is adopted 
$$
\begin{aligned}
\begin{matrix}
\boldsymbol{\beta}^{(i+1)} = \boldsymbol{\beta}^{(i)}- S_{\boldsymbol{\beta}}^{-1} \psi \boldsymbol{\beta} (\boldsymbol{\beta}^{(i)}, \boldsymbol{\lambda}^{(i)}), \\ 
\boldsymbol{\lambda}^{(i+1)} = \boldsymbol{\lambda}^{(i)}\alpha S_{\boldsymbol{\lambda}}^{-1} \psi \boldsymbol{\lambda} (\boldsymbol{\beta}^{(i+1)}, \boldsymbol{\lambda}^{(i)}).
\end{matrix}
\end{aligned}
$$

The procedure aforementioned is implements in the \pkg{mcglm} package \citep{mcglm}.

@@

Following the ideas of \textbf{REF MEU ARTIGO}, let $\boldsymbol{\theta^{*}}$ be a $h \times 1$ vector of parameters disregarding the correlation parameters and $J^{\boldsymbol{*}-1}$ the corresponding inverse of the Godambe information matrix. Let $\boldsymbol{L}$ be a $s \times h$ hypotheses specification matrix and $\boldsymbol{c}$ a $s \times 1$ vector with the values under the null hypothesis. In this notation, $s$ represents the number of restrictions. The hypotheses to be tested can be written as:

@@

\noindent Thus, the generalization of the Wald test statistic to verify the validity of a hypothesis about parameters of a McGLM is given by:

@@

In general, each column of the matrix $\boldsymbol{L}$ corresponds to one of the $h$ parameters of $\boldsymbol{\theta^{*}}$ and each row to a constraint. Its specification basically consists of filling the matrix with $0$, $1$ and eventually $-1$ in such a way that the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ correctly represents the hypotheses of interest. The correct specification of $\boldsymbol{L}$ allows us testing any linear hipothesis for each parameter individually or even formulating hypotheses for several parameters jointly.

\textbf{REF MEU ARTIGO} presents examples of how to test different types of hypotheses of interest that arise in practical contexts. In this article, we shall present two examples: hypotheses for multiple parameters and hypotheses considering regression or dispersion parameters for responses under the same linear predictor.

For purposes of illustration, consider the case in which one wants to investigate whether a numeric variable $X_1$ has an effect on two response variables. Let $Y_1$ and $Y_2$ denote the response variables. A bivariate McGLM for this problem may have a linear predictor given by:

@@

\noindent where the index $r$ denotes the response variable, $r = 1,2$; $\beta_{r0}$ represents the intercept; $\beta_{r1}$ the regression coefficient associated with the variable $X_1$. We assume that each response has only one dispersion parameter $\tau_{r0}$ and that the power parameters were fixed. Therefore, it is a problem in which there are two response variables and only one explanatory variable. Further, we assume that the observations are independent, so $Z_0 = I$.

Suppose the interest is to assess whether there is sufficient evidence to state that there is an effect of the explanatory variable $X_1$ on both response variables simultaneously. 
In this cas, we  have to test two parameters: $\beta_{11}$, which associates $X_1$ with the first response variable; and $\beta_{21}$, which associates $X_1$ with the second response variable. 
We can write the hypothesis as follows:

@@

The vector $\boldsymbol{\theta^{*}}$ has six elements and the matrix $\boldsymbol{L}$ has six columns. In this case, we are testing two parameters, so the matrix $\boldsymbol{L}$ has two rows. These lines are composed of zeros, except in the columns referring to the parameter of interest. It is simple to verify that the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ represents the hypothesis of interest, see Eq. \autoref{eq:ex2}. 
Thus, the asymptotic distribution of the test is $\chi^2_2$.

@@

Suppose that, in this example, the hypotheses of interest are still written as in the form of \autoref{eq:ex2}. However, as this is a bivariate model with the same predictor for the two responses, the hypothesis of interest are the same between responses and involves only regression parameters. Consequently, it is convenient to write the matrix $\boldsymbol{L}$ as the Kronecker product of two matrices: a matrix $\boldsymbol{G}$ and a matrix $\boldsymbol{F}$, ie, $\boldsymbol{L}$ = $\boldsymbol{G} \otimes \boldsymbol{F}$. In this way, the matrix $\boldsymbol{G}$ has dimension $R \times R$ and specifies the hypotheses about the responses, whereas the matrix $\boldsymbol{F}$ specifies the hypotheses between variables and has dimension ${s}' \times {h}'$, where ${s}'$ is the number of linear constraints, that is, the number of parameters tested for a single response, and ${h}'$ is the total number of coefficients of regression or dispersion of the response. Therefore, the matrix $\boldsymbol{L}$ has dimension (${s}'R \times h$).

In general, the matrix $\boldsymbol{G}$ is an identity matrix with a dimension equal to the number of responses composing the model. Whereas the matrix $\boldsymbol{F}$ is equivalent to a matrix $\boldsymbol{L}$ if there was only a single response in the model and only regression or dispersion parameters. We use the Kronecker product of these two matrices to ensure that the hypothesis described in the $\boldsymbol{F}$ matrix will be tested on $R$ model responses.

@@

Based on the Wald statistics adapted for McGLMs, \textbf{REF MEU ARTIGO} proposed three different procedures for generating ANOVA and MANOVA tables for regression parameters, and a procedure similar to ANOVA and MANOVA to evaluate the dispersion parameters of a model. In the case of ANOVAs, a table is generated for each response variable. For MANOVAs only one table is generated, therefore, in order to be able to perform MANOVAs, the responses must be subject to the same linear predictor.

For purposes of illustration, consider the situation where the goal is to investigate whether two numeric variables denoted by $X_1$ and $X_2$ have an effect on two response variables denoted by $Y_1$ and $Y_2$. For this case, consider the following linear predictor:

@@

In this section we shall provide practical examples of using the functions implemented in the \pkg{htmcglm} package based on multivariate models fitted with the \pkg{mcglm} package.
